---
title: "FaeSeq-E745 - Differential Expression Analysis"
author: "Victor Cornejo Villanueva"
date: "2026-05-11"
output: html_document
---

# Load the packages
```{r}
library("DESeq2")
library("ggplot2")
library("ggrepel")
```

# 1. Setup the files path
```{r}
directory <- "~/Escritorio/Uppsala/9) Genome Analysis/Project/DESeq"
```

# 2. Build the Metadata Table
```{r}
sample_names <- c("ERR1797969", "ERR1797970", "ERR1797971", "ERR1797972", "ERR1797973", "ERR1797974")
conditions <- c("Serum", "Serum", "Serum", "BH", "BH", "BH")
file_names <- paste0(sample_names, "_counts.txt")

sampleTable <- data.frame(sampleName = sample_names,
                          fileName = file_names,
                          condition = factor(conditions))
```

# 3. Import the HTSeq data directly into a DESeq2 object
```{r}
dds <- DESeqDataSetFromHTSeqCount(sampleTable = sampleTable,
                                  directory = directory,
                                  design = ~ condition)
```

# 4. Filter out genes that have zero or almost zero expression across all samples
```{r}
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```

# 5. Run the actual Differential Expression Analysis
```{r}
dds <- DESeq(dds)
```

# Check the results summary
```{r}
res <- results(dds)
summary(res)
```

# Extract the list of genes
```{r}
# Sort the results by the adjusted p-value (most significant at the top)
resOrdered <- res[order(res$padj),]

# View the top 10 genes directly in R
head(resOrdered, 10)

# Export the entire list to a CSV file on your computer
write.csv(as.data.frame(resOrdered), file="E745_Serum_vs_BH_results.csv")
```

# Associate the list of genes with annotation file
```{r}
# Load the Prokka dictionary
anno <- read.delim("DESeq/e745_annot.tsv", stringsAsFactors = FALSE)

# Convert the DESeq2 results into a standard dataframe
res_df <- as.data.frame(resOrdered)

# Create a locus_tag column from the row names
res_df$locus_tag <- rownames(res_df)

# Merge the two tables together based on the locus_tag
final_results <- merge(res_df, anno, by="locus_tag", all.x=TRUE)

# Sort the final merged table by significance again
final_results <- final_results[order(final_results$log2FoldChange),]

# Peek at the top 10 genes with their actual biological names!
head(final_results[, c("locus_tag", "log2FoldChange", "padj", "gene", "product")], 10)

# Export the ultimate master spreadsheet
write.csv(final_results, file="E745_Serum_vs_BH_Master_Annotated.csv", row.names=FALSE)
```

# Principal Component Analysis (PCA)
```{r}
# Normalize the variance so the plot isn't skewed by highly expressed genes
vsd <- vst(dds, blind=FALSE)

# Extract the raw data
pcaData <- plotPCA(vsd, intgroup="condition", returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

# Build the plot 
pca_plot <- ggplot(pcaData, aes(x=PC1, y=PC2, color=condition)) +
  geom_point(size=6) +          
  scale_color_manual(values=c("blue", "red")) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) +
  theme_minimal() +
  ggtitle("PCA - Serum vs BH Replicates") +
  theme(plot.title = element_text(hjust = 0.5, face="bold", size=16),
        legend.title = element_blank(),
        legend.text = element_text(size=12))
pca_plot
```

# Volcano Plot
```{r}
plot_data <- final_results
plot_data <- plot_data[!is.na(plot_data$gene) & plot_data$gene != "", ]

plot_data$Significance <- "Not Significant"
plot_data$Significance[plot_data$log2FoldChange > 1 & plot_data$padj < 0.05] <- "Up in Serum"
plot_data$Significance[plot_data$log2FoldChange < -1 & plot_data$padj < 0.05] <- "Down in Serum"

# Label Logic
plot_data$label <- NA
# Threshold > 4
extreme_genes <- abs(plot_data$log2FoldChange) > 4
plot_data$label[extreme_genes] <- plot_data$gene[extreme_genes]

# Draw the Volcano Plot
volcano_plot <- ggplot(plot_data, aes(x=log2FoldChange, y=-log10(padj), color=Significance)) +
  geom_point(alpha=0.8, size=2) +
  scale_color_manual(values=c("blue", "grey", "red")) +
  theme_minimal() +
  geom_vline(xintercept=c(-1, 1), linetype="dashed", color="black") +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color="black") +
  geom_text_repel(aes(label=label), 
                  color="black",
                  max.overlaps=Inf,
                  box.padding=0.5,    
                  point.padding=0.3,  
                  size=3.5, 
                  fontface="italic") +
  labs(title="Volcano Plot - Transcriptomic Response to Serum", 
       x="Log2 Fold Change", 
       y="-Log10 Adjusted P-value") +
  theme(legend.position="bottom", 
        plot.title = element_text(hjust = 0.5, face="bold", size=16),
        axis.text = element_text(size=12),
        axis.title = element_text(size=14))

# Save the plot
ggsave("E745_Volcano_Clean.png", plot=volcano_plot, width=13, height=12, dpi=300)

# Print to screen
print(volcano_plot)
```
